---
title: "Coursera: Practical Machin Learning"
output: html_notebook
---

__Author:__ Seth Dimick  
__Date:__ `r format(Sys.Date(), '%m/%d/%Y')`  
__Repo:__ https://github.com/sdimick/coursera-pml  

### Objective

The final assignment of this [Practical Machine Learning course](https://www.coursera.org/learn/practical-machine-learning/home/welcome) is to build a classification model. The model is based off exercise accelerometer data from Groupware\@LES' [Human Activity Recognition](http://groupware.les.inf.puc-rio.br/har) project, where participants were instructed to perform different lifts correctly and incorrectly in five distinct ways. Trained on a predefined training data set, the model should be able to accurately predict which of the five ways a lift was performed on a 20 observation testing data set.

### Get the Data

For the sake of reproducibility, I have included the code to download the data sets and clean up the training data for modeling.

```{r}
# Training data set
trainingURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
trainingPath <- "../data/pml-training.csv"
download.file(trainingURL, trainingPath, quiet = TRUE)
trainingDF <- read.csv(trainingPath)

# Testing data set
testingURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
testingPath <- "../data/pml-testing.csv"
download.file(testingURL, testingPath, quiet = TRUE)
testingDF <- read.csv(testingPath)
```

Upon download, you will notice that many of the 160 columns in the training data set are full of blanks and/or NAs. Next I use the following script to subset down to the usable variables based on availability.

```{r}
# Keep columns with more than 90% of rows populated
keepCols <- numeric()
for (i in 1:ncol(trainingDF)) {
	testCol <- trainingDF[, i]
	keepCondition <- ifelse(
		sum(is.na(testCol) | testCol == "") > 0.9 * length(testCol),
		FALSE,
		TRUE
	)
	if (keepCondition) {
		keepCols <- c(keepCols, i)
	}
}
trainingDF <- trainingDF[, keepCols]
```



### Modeling

blah

### Results

blah
